### 2025-03-15 paper/quarto_ecml/paper.tex
Back to 16 pages, emphasize robustness a bit more

The first challenge has attracted an abundance of research on
\textbf{explainable AI} (XAI), a paradigm that focuses on the
development of tools to derive (post-hoc) explanations from complex
model representations. Such explanations should mitigate a scenario in
which practitioners deploy opaque models and blindly rely on their
predictions. On countless occasions, this has happened in practice and
caused real harms to people who were adversely and unfairly affected by
automated decision-making (ADM) systems involving opaque models
\cite{mcgregor2021preventing,oneil2016weapons}. Effective XAI tools can
aid us in monitoring models and providing recourse to individuals to
turn negative outcomes (e.g., ``loan application rejected'') into
positive ones (e.g., ``application accepted''). In line with this, our
work builds upon \textbf{counterfactual explanations} (CE) proposed by
\cite{wachter2017counterfactual} as an effective approach to achieve
this goal. CEs prescribe minimal changes for factual inputs that, if
implemented, would prompt some fitted model to produce a desired output.

@@

The remainder of this manuscript is structured as follows.
Section~\ref{sec-lit} presents related work, focusing on the link
between AEs and CEs. Then follow our two principal contributions. In
Section~\ref{sec-method}, we introduce our methodological framework and
show theoretically that it can be employed to enforce global
actionability constraints. In Section~\ref{sec-experiments}, through
extensive experiments, we demonstrate that CT substantially improves
explainability without sacrificing predictive performance. We discuss
the challenges in Section~\ref{sec-discussion} and conclude in
Section~\ref{sec-conclusion} that CT is a promising approach towards
making opaque models more trustworthy.

@@

\caption{\label{fig-poc}Illustration of how CT improves model
explainability: (a) conventional training, all mutable; (b) CT, all
mutable; (c) conventional, \emph{age} immutable; (d) CT, \emph{age}
immutable. The decision boundary is shown in green along with training
data colored according to their ground-truth label: \(y^+=2\) (orange)
and \(y^-=1\) (blue). Stars indicate CEs for the target class.}

@@

Our key outcome of interest is improved explainability (Def.
\ref{def-explainability}). To this end, we focus primarily on the
plausibility and cost of faithfully generated counterfactuals at test
time. To measure the cost of counterfactuals, we follow the standard
convention of using distances (\(\ell_1\)-norm) between factuals and
counterfactuals as a proxy. For plausibility, we assess how similar
counterfactuals are to observed samples in the target domain,
\(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the distance-based
metric used in \cite{altmeyer2024faithful},

@@

To assess our proposed training regime, we investigate the improvements
in performance metrics when using CT on top of a weak baseline (BL):
specifically, a multilayer perceptron (\emph{MLP}). This is the best way
to get a clear picture of how effective CT is and consistent with how
assessment is done in the related literature
\cite{ross2021learning,teney2020learning,goodfellow2014explaining}.

@@

Table~\ref{tbl-main} presents our main empirical findings. For all
datasets except \emph{OL} and across all test settings, the average
distance of CEs from observed samples in the target class is reduced,
indicating improved plausibility. The magnitude of improvements varies.
For the simple synthetic datasets, distance reductions range from around
20-40\% (\emph{LS}, \emph{Moon}) to almost 60\% (\emph{Circ}). For the
real-world tabular datasets, improvements tend to be smaller but still
substantial, with around 10-15\% for \emph{CH}, 11-28\% for \emph{GMSC},
7-8\% for \emph{Cred}, and around 3\% for \emph{Adult}. For our only
vision dataset (\emph{MNIST}), distances are reduced by up to 9\%. The
results for our proposed divergence metric are qualitatively similar,
but generally even more pronounced: for the \emph{Circ} dataset,
implausibility is reduced by almost 94\% to virtually zero as we
verified by the absolute outcome. Improvements for other datasets range
from 28\% (\emph{Moon}) to 78\% (\emph{GMSC}). For \emph{OL} the
reduction is negative, consistent with the distance-based metric.
\emph{MNIST} is the only dataset for which the distance and divergence
metrics disagree. Upon visual inspection of the image counterfactuals we
find that CT clearly improves plausibility (see appendix).

@@

\caption{\label{tbl-main}Key performance metrics across all datasets.
\textbf{Plausibility}: Columns 2-6 show the percentage reduction in
implausibility (\(\text{IP}\)) for varying degrees of the energy penalty
used for \emph{ECCo} at test time; column 7 shows the reduction in
\(\text{IP}^*\) (\(\text{MMD}\)), aggregated across all test
specifications. \textbf{Accuracy} (columns 8-11): test accuracies and
robust accuracies (\(\text{Acc}^*\)) for CT and the baseline (BL).
\textbf{Actionability} (column 9): average reduction in costs when
imposing mutability constraints.}

@@

\section{Discussion}\label{sec-discussion}

As our results indicate, counterfactual training produces models that
are more explainable. Nonetheless, it brings about three important
limitations.\\
\indent \emph{CT increases the training time of models.} CT can be more
time-consuming than conventional training regimes. While higher numbers
of CEs per iteration positively impact the quality of solutions, they
also increase the amount of computations. Relatively small grids with
270 settings can take almost four hours for more demanding datasets on a
high-performance computing cluster with 34 2GB CPUs.\footnote{See
  supplementary appendix for computational details.} Three factors
attenuate this effect. First, CT amortizes the cost of CEs for the
training samples. Second, we find that it can retain its value when used
as a ``fine-tuning'' technique for conventionally-trained models. Third,
it yields itself to parallel execution, which we have leveraged for our
own experiments.\\
\indent \emph{Immutable features may have proxies.} We propose an
approach to protect immutable features and thus increase the
actionability of the generated CEs. However, it requires that model
owners define the mutability constraints for (all) features considered
by the model. Even if all immutable features are protected, there may
exist proxies that are mutable (and hence should not be protected) but
preserve enough information about the principals to hinder the
protections. Delineating actionability is a major undecided challenge in
the AR literature (see, e.g.,
\cite{venkatasubramanian2020philosophical}) impacting the capacity of CT
to fulfill its intended goal.\\
\indent \emph{Interventions on features may impact fairness.} We provide
a tool that allows practitioners to modify the sensitivity of a model
with respect to certain features, which may have implication for the
fair and equitable treatment of decision subjects. As protecting a set
of features leads the model to assign higher relative importance to
unprotected features, model owners could misuse our solution by
enforcing explanations based on features that are more difficult to
modify by some (group of) individuals. For example, consider the Adult
dataset used in our experiments, where \emph{workclass} or
\emph{education} may be more difficult to change for underpriviledged
groups. When applied irresponsibly, CT could result in an unfairly
assigned burden of recourse \cite{sharma2020certifai}, threatening the
equality of opportunity in the system \cite{bell2024fairness}.
Nonetheless, these phenomena are not specific to CT.

We also highlight several important directions for future research.
Firstly, it is an interesting challenge to extend CT beyond
classification settings. Our formulation relies on the distinction
between non-target class(es) \(y^{-}\) and target class(es) \(y^{+}\) to
generate counterfactuals through Equation~\ref{eq-obj}. While \(y^{-}\)
and \(y^{+}\) can be arbitrarily defined, CT requires the output space
\(\mathcal{Y}\) to be discrete. Thus, it does not apply to ML tasks
where the change in outcome cannot be readily quantified. Focus on
classification models is a common restriction in research on CEs and AR.
Other settings have attracted some interest (e.g., regression in
\cite{spooner2021counterfactual}), but there is little consensus how to
robustly extend the notion of CEs.\\
\indent Secondly, our approach is susceptible to training instabilities.
This problem has been recognized for JEMs \cite{grathwohl2020your} and
even though we depart from the SGLD-based sampling, we still encounter
considerable variability in the outcomes. CT is exposed to two potential
sources of instabilities: (1) the energy-based contrastive divergence
term in Equation~\ref{eq-div}, and (2) the underlying counterfactual
explainers. We find several promising ways to mitigate this problem:
regularizing energy (\(\lambda_{\text{reg}}\)), generating sufficiently
many counterfactuals during each epoch, and including only mature
counterfactuals for contrastive divergence.\\
\indent Finally, we believe that it is possible to substantially improve
hyperparameter selection procedures. Our method benefits from the tuning
of certain key hyperparameters (see Section~\ref{sec-hyperparameters}).
In this work, we have relied exclusively on grid search for this task.
Future work on CT could benefit from investigating more sophisticated
approaches. Notably, CT is iterative which makes methods such as
Bayesian or gradient-based optimization applicable (see, e.g.,
\cite{bischl2023hyperparameter}).

\section{Conclusion}\label{sec-conclusion}

State-of-the-art machine learning models are prone to learning complex
representations that cannot be interpreted by humans and existing
post-hoc explainability approaches cannot guarantee that the
explanations agree with the model's learned representation of data. As a
step towards addressing this challenge, we introduced counterfactual
training, a novel training regime that incentivizes highly-explainable
models. Our approach leads to explanations that are both
plausible---compliant with the underlying data-generating process---and
actionable---compliant with user-specified mutability constraints---and
thus meaningful to their recipients. Through extensive experiments we
demonstrate that CT satisfies its objectives while preserving the
predictive performance of the models. Our approach can also be used to
fine-tune conventionally-trained models and achieve similar gains in
explainability. Finally, this work showcases that it is practical to
improve models \emph{and} their explanations at the same time.

>>>>

The first challenge has attracted an abundance of research on
\textbf{explainable AI} (XAI), a paradigm that focuses on the
development of tools to derive (post-hoc) explanations from complex
model representations. Such explanations should mitigate a scenario in
which practitioners deploy opaque models and blindly rely on their
predictions. On countless occasions, this has happened in practice and
caused real harms to people who were adversely and unfairly affected by
automated decision-making (ADM) systems involving opaque models (see,
e.g., \cite{oneil2016weapons}). Effective XAI tools can aid us in
monitoring models and providing recourse to individuals to turn negative
outcomes (e.g., ``loan application rejected'') into positive ones (e.g.,
``application accepted''). In line with this, our work builds upon
\textbf{counterfactual explanations} (CE) proposed by
\cite{wachter2017counterfactual} as an effective approach to achieve
this goal. CEs prescribe minimal changes for factual inputs that, if
implemented, would prompt some fitted model to produce a desired output.

@@

The remainder of this manuscript is structured as follows.
Section~\ref{sec-lit} presents related work, focusing on the link
between AEs and CEs. Then follow our two principal contributions. In
Section~\ref{sec-method}, we introduce our methodological framework and
show theoretically that it can be employed to enforce global
actionability constraints. In Section~\ref{sec-experiments}, through
extensive experiments, we demonstrate that CT substantially improves
explainability and positively contributes to the robustness of trained
models without sacrificing predictive performance. Finally, in
Section~\ref{sec-discussion}, we discuss the challenges and conclude
that CT is a promising approach towards making opaque models more
trustworthy.

@@

\caption{\label{fig-poc}Illustration of how CT improves model
explainability: (a) conventional training, all mutable; (b) CT, all
mutable; (c) conventional, \emph{age} immutable; (d) CT, \emph{age}
immutable. The linear decision boundary is shown in green along with
training data colored according to their ground-truth label: \(y^-=1\)
in blue and \(y^+=2\) in orange. Stars indicate counterfactuals in the
target class.}

@@

Our key outcome of interest is improvement in explainability (Def.
\ref{def-explainability}). To this end, we focus primarily on the
plausibility and cost of faithfully generated counterfactuals at test
time. To measure the cost, we follow the standard convention of using
distances (\(\ell_1\)-norm) between factuals and counterfactuals as a
proxy. For plausibility, we assess how similar CEs are to the observed
samples in the target domain, \(\mathbf{X}^+\subset\mathcal{X}^+\). We
rely on the distance-based metric used in \cite{altmeyer2024faithful},

@@

To assess CT, we investigate the improvements in performance metrics
when using it on top of a weak baseline (BL): a multilayer perceptron
(\emph{MLP}). This is the best way to get a clear picture of the
effectiveness of CT, and it is consistent with how assessment is done in
the related literature
\cite{goodfellow2014explaining,ross2021learning,teney2020learning}.

@@

Table~\ref{tbl-main} presents our main empirical findings. For all
datasets except \emph{OL} and across all test settings, the average
distance of CEs from observed samples in the target class is reduced,
indicating improved plausibility. The magnitude of improvements varies.
For the simple synthetic datasets, distance reductions range from around
20-40\% (\emph{LS}, \emph{Moon}) to almost 60\% (\emph{Circ}). For the
real-world tabular datasets, improvements tend to be smaller but still
substantial, with around 10-15\% for \emph{CH}, 11-28\% for \emph{GMSC},
7-8\% for \emph{Cred}, and around 3\% for \emph{Adult}. For the vision
dataset (\emph{MNIST}), distances are reduced by up to 9\%. The results
for our proposed divergence metric are qualitatively similar, but
generally even more pronounced: for the \emph{Circ} dataset,
implausibility is reduced by almost 94\% to virtually zero as we
verified by the absolute outcome. Improvements for other datasets range
from 28\% (\emph{Moon}) up to 78\% (\emph{GMSC}). For \emph{OL} the
reduction is negative, consistent with the distance-based metric.
\emph{MNIST} is the only dataset for which the distance and divergence
metrics disagree. Upon visual inspection of the image counterfactuals we
find that CT clearly improves plausibility (see supplementary appendix).

@@

\caption{\label{tbl-main}Key performance metrics across all datasets
(column 1). \textbf{Plausibility}: Columns 2-6 show the percentage
reduction in implausibility (\(\text{IP}\)) for varying degrees of the
energy penalty \(\lambda_{\text{egy}}\) used for \emph{ECCo} at test
time; column 7 shows the reduction in \(\text{IP}^*\) (\(\text{MMD}\)),
aggregated across all test specifications. \textbf{Accuracy} (columns
8-11): test accuracies and robust accuracies (\(\text{Acc}^*\)) for CT
and the baseline (BL). \textbf{Actionability} (column 12): average
reduction in costs when imposing mutability constraints reported for the
four datasets for which we could identify meaningful features to
protect.}

@@

\section{Conclusions}\label{sec-discussion}

As our results indicate, counterfactual training produces models that
are more explainable. Nonetheless, it brings about three important
limitations.\\
\indent \emph{CT increases the training time of models.} CT can be more
time-consuming than conventional training regimes. While higher numbers
of CEs per iteration positively impact the quality of solutions, they
also increase the amount of computations. Relatively small grids with
270 settings can take almost four hours for more demanding datasets on a
high-performance computing cluster with 34 2GB CPUs.\footnote{See
  supplementary appendix for computational details.} Three factors
attenuate this effect: (1) CT amortizes the cost of CEs for the training
samples; (2) it can retain its value when used as a ``fine-tuning''
technique for conventionally-trained models; and (3) it yields itself to
parallel execution, which we have leveraged for our own experiments.\\
\indent \emph{Immutable features may have proxies.} We propose an
approach to protect immutable features and thus increase the
actionability of the generated CEs. However, it requires that model
owners define the mutability constraints for (all) features considered
by the model. Even if all immutable features are protected, there may
exist proxies that are mutable (and hence should not be protected) but
preserve enough information about the principals to hinder the
protections. Delineating actionability is a major open challenge in the
AR literature (see, e.g., \cite{venkatasubramanian2020philosophical})
impacting the capacity of CT to fulfill its intended goal.\\
\indent \emph{Interventions on features may impact fairness.} We provide
a tool that allows practitioners to modify the sensitivity of a model
with respect to certain features, which may have implication for the
fair and equitable treatment of decision subjects. Model owners could
misuse this solution by enforcing explanations based on features that
are more difficult to modify by some (group of) individuals. For
example, consider the \emph{Adult} dataset used in our experiments,
where \emph{workclass} or \emph{education} may be more difficult to
change for underpriviledged groups. When applied irresponsibly, CT could
result in an unfairly assigned burden of recourse
\cite{sharma2020certifai}, threatening the equality of opportunity in
the system \cite{bell2024fairness}. Nonetheless, these phenomena are not
specific to CT.\\
\hfill \break \indent We also highlight several important directions for
future research. Firstly, it is an interesting challenge to extend CT
beyond classification settings. Our formulation relies on the
distinction between non-target class(es) \(y^{-}\) and target class(es)
\(y^{+}\) to generate counterfactuals through Equation~\ref{eq-obj}.
While \(y^{-}\) and \(y^{+}\) can be arbitrarily defined, CT requires
the output space \(\mathcal{Y}\) to be discrete. Thus, it does not apply
to ML tasks where the change in outcome cannot be readily quantified.
Focus on classification models is a common restriction in research on
CEs and AR. Other settings have attracted some interest (e.g.,
regression in \cite{spooner2021counterfactual}), but there is little
consensus how to robustly extend the notion of CEs.\\
\indent Secondly, our approach is susceptible to training instabilities.
This problem has been recognized for JEMs \cite{grathwohl2020your} and
even though we depart from the SGLD-based sampling, we still encounter
considerable variability in the outcomes. CT is exposed to two potential
sources of instabilities: (1) the energy-based contrastive divergence
term in Equation~\ref{eq-div}, and (2) the underlying counterfactual
explainers. We find several promising ways to mitigate this problem:
regularizing energy (\(\lambda_{\text{reg}}\)), generating sufficiently
many counterfactuals during each epoch, and including only mature
counterfactuals for contrastive divergence.\\
\indent Finally, we believe that it is possible to substantially improve
hyperparameter selection procedures. Our method benefits from the tuning
of certain key hyperparameters (see Section~\ref{sec-hyperparameters}).
In this work, we have relied exclusively on grid search for this task.
Future work on CT could benefit from investigating more sophisticated
approaches. Notably, CT is iterative which makes methods such as
Bayesian or gradient-based optimization applicable (see, e.g.,
\cite{bischl2023hyperparameter}). \hfill \break

To conclude, state-of-the-art machine learning models are prone to
learning complex representations that cannot be interpreted by humans
and existing post-hoc explainability approaches cannot guarantee that
the explanations agree with the model's learned representation of data.
As a step towards addressing this challenge, we introduced
counterfactual training, a novel training regime that incentivizes
highly-explainable models. Our approach leads to explanations that are
both plausible---compliant with the underlying data-generating
process---and actionable---compliant with user-specified mutability
constraints---and thus meaningful to their recipients. Through extensive
experiments we demonstrate that CT satisfies its objective while
promoting robustness and preserving the predictive performance of the
models. It can also be used to fine-tune conventionally-trained models
and achieve similar gains. Finally, this work showcases that it is
practical to improve models \emph{and} their explanations at the same
time.

### 2025-03-15 paper/sections/conclusion.qmd
Back to 16 pages, emphasize robustness a bit more

# Conclusion {#sec-conclusion}
State-of-the-art machine learning models are prone to learning complex representations that cannot be interpreted by humans and existing post-hoc explainability approaches cannot guarantee that the explanations agree with the model's learned representation of data. As a step towards addressing this challenge, we introduced counterfactual training, a novel training regime that incentivizes highly-explainable models. Our approach leads to explanations that are both plausible---compliant with the underlying data-generating process---and actionable---compliant with user-specified mutability constraints---and thus meaningful to their recipients. Through extensive experiments we demonstrate that CT satisfies its objectives while preserving the predictive performance of the models. Our approach can also be used to fine-tune conventionally-trained models and achieve similar gains in explainability. Finally, this work showcases that it is practical to improve models *and* their explanations at the same time.

>>>>

<!-- # Conclusion {#sec-conclusion} -->
To conclude, state-of-the-art machine learning models are prone to learning complex representations that cannot be interpreted by humans and existing post-hoc explainability approaches cannot guarantee that the explanations agree with the model's learned representation of data. As a step towards addressing this challenge, we introduced counterfactual training, a novel training regime that incentivizes highly-explainable models. Our approach leads to explanations that are both plausible---compliant with the underlying data-generating process---and actionable---compliant with user-specified mutability constraints---and thus meaningful to their recipients. Through extensive experiments we demonstrate that CT satisfies its objective while promoting robustness and preserving the predictive performance of the models. It can also be used to fine-tune conventionally-trained models and achieve similar gains. Finally, this work showcases that it is practical to improve models *and* their explanations at the same time.

### 2025-03-15 paper/sections/discussion.qmd
Back to 16 pages, emphasize robustness a bit more

# Discussion {#sec-discussion}

As our results indicate, counterfactual training produces models that are more explainable. Nonetheless, it brings about three important limitations.  
`\indent`{=latex} *CT increases the training time of models.* CT can be more time-consuming than conventional training regimes. While higher numbers of CEs per iteration positively impact the quality of solutions, they also increase the amount of computations. Relatively small grids with 270 settings can take almost four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs.^[See supplementary appendix for computational details.] Three factors attenuate this effect. First, CT amortizes the cost of CEs for the training samples. Second, we find that it can retain its value when used as a "fine-tuning" technique for conventionally-trained models. Third, it yields itself to parallel execution, which we have leveraged for our own experiments.  
`\indent`{=latex} *Immutable features may have proxies.* We propose an approach to protect immutable features and thus increase the actionability of the generated CEs. However, it requires that model owners define the mutability constraints for (all) features considered by the model. Even if all immutable features are protected, there may exist proxies that are mutable (and hence should not be protected) but preserve enough information about the principals to hinder the protections. Delineating actionability is a major undecided challenge in the AR literature [see, e.g., @venkatasubramanian2020philosophical] impacting the capacity of CT to fulfill its intended goal.  
`\indent`{=latex} *Interventions on features may impact fairness.* We provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features, which may have implication for the fair and equitable treatment of decision subjects. As protecting a set of features leads the model to assign higher relative importance to unprotected features, model owners could misuse our solution by enforcing explanations based on features that are more difficult to modify by some (group of) individuals. For example, consider the Adult dataset used in our experiments, where *workclass* or *education* may be more difficult to change for underpriviledged groups. When applied irresponsibly, CT could result in an unfairly assigned burden of recourse [@sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness]. Nonetheless, these phenomena are not specific to CT.

We also highlight several important directions for future research. Firstly, it is an interesting challenge to extend CT beyond classification settings. Our formulation relies on the distinction between non-target class(es) $y^{-}$ and target class(es) $y^{+}$ to generate counterfactuals through @eq-obj. While $y^{-}$ and $y^{+}$ can be arbitrarily defined, CT requires the output space $\mathcal{Y}$ to be discrete. Thus, it does not apply to ML tasks where the change in outcome cannot be readily quantified. Focus on classification models is a common restriction in research on CEs and AR. Other settings have attracted some interest (e.g., regression in [@spooner2021counterfactual]), but there is little consensus how to robustly extend the notion of CEs.  
`\indent`{=latex} Secondly, our approach is susceptible to training instabilities. This problem has been recognized for JEMs [@grathwohl2020your] and even though we depart from the SGLD-based sampling, we still encounter considerable variability in the outcomes. CT is exposed to two potential sources of instabilities: (1) the energy-based contrastive divergence term in @eq-div, and (2) the underlying counterfactual explainers. We find several promising ways to mitigate this problem: regularizing energy ($\lambda_{\text{reg}}$), generating sufficiently many counterfactuals during each epoch, and including only mature counterfactuals for contrastive divergence.  
`\indent`{=latex} Finally, we believe that it is possible to substantially improve hyperparameter selection procedures. Our method benefits from the tuning of certain key hyperparameters (see @sec-hyperparameters). In this work, we have relied exclusively on grid search for this task. Future work on CT could benefit from investigating more sophisticated approaches. Notably, CT is iterative which makes methods such as Bayesian or gradient-based optimization applicable (see, e.g., [@bischl2023hyperparameter]).

>>>>

# Conclusions {#sec-discussion}

As our results indicate, counterfactual training produces models that are more explainable. Nonetheless, it brings about three important limitations.  
`\indent`{=latex} *CT increases the training time of models.* CT can be more time-consuming than conventional training regimes. While higher numbers of CEs per iteration positively impact the quality of solutions, they also increase the amount of computations. Relatively small grids with 270 settings can take almost four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs.^[See supplementary appendix for computational details.] Three factors attenuate this effect: (1) CT amortizes the cost of CEs for the training samples; (2) it can retain its value when used as a "fine-tuning" technique for conventionally-trained models; and (3) it yields itself to parallel execution, which we have leveraged for our own experiments.  
`\indent`{=latex} *Immutable features may have proxies.* We propose an approach to protect immutable features and thus increase the actionability of the generated CEs. However, it requires that model owners define the mutability constraints for (all) features considered by the model. Even if all immutable features are protected, there may exist proxies that are mutable (and hence should not be protected) but preserve enough information about the principals to hinder the protections. Delineating actionability is a major open challenge in the AR literature (see, e.g., [@venkatasubramanian2020philosophical]) impacting the capacity of CT to fulfill its intended goal.  
`\indent`{=latex} *Interventions on features may impact fairness.* We provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features, which may have implication for the fair and equitable treatment of decision subjects. Model owners could misuse this solution by enforcing explanations based on features that are more difficult to modify by some (group of) individuals. For example, consider the *Adult* dataset used in our experiments, where *workclass* or *education* may be more difficult to change for underpriviledged groups. When applied irresponsibly, CT could result in an unfairly assigned burden of recourse [@sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness]. Nonetheless, these phenomena are not specific to CT.  
`\hfill \break`{=latex}
`\indent`{=latex} We also highlight several important directions for future research. Firstly, it is an interesting challenge to extend CT beyond classification settings. Our formulation relies on the distinction between non-target class(es) $y^{-}$ and target class(es) $y^{+}$ to generate counterfactuals through @eq-obj. While $y^{-}$ and $y^{+}$ can be arbitrarily defined, CT requires the output space $\mathcal{Y}$ to be discrete. Thus, it does not apply to ML tasks where the change in outcome cannot be readily quantified. Focus on classification models is a common restriction in research on CEs and AR. Other settings have attracted some interest (e.g., regression in [@spooner2021counterfactual]), but there is little consensus how to robustly extend the notion of CEs.  
`\indent`{=latex} Secondly, our approach is susceptible to training instabilities. This problem has been recognized for JEMs [@grathwohl2020your] and even though we depart from the SGLD-based sampling, we still encounter considerable variability in the outcomes. CT is exposed to two potential sources of instabilities: (1) the energy-based contrastive divergence term in @eq-div, and (2) the underlying counterfactual explainers. We find several promising ways to mitigate this problem: regularizing energy ($\lambda_{\text{reg}}$), generating sufficiently many counterfactuals during each epoch, and including only mature counterfactuals for contrastive divergence.  
`\indent`{=latex} Finally, we believe that it is possible to substantially improve hyperparameter selection procedures. Our method benefits from the tuning of certain key hyperparameters (see @sec-hyperparameters). In this work, we have relied exclusively on grid search for this task. Future work on CT could benefit from investigating more sophisticated approaches. Notably, CT is iterative which makes methods such as Bayesian or gradient-based optimization applicable (see, e.g., [@bischl2023hyperparameter]).
`\hfill \break`{=latex}

### 2025-03-15 paper/sections/experiments.qmd
Back to 16 pages, emphasize robustness a bit more

Our key outcome of interest is improved explainability (Def. \ref{def-explainability}). To this end, we focus primarily on the plausibility and cost of faithfully generated counterfactuals at test time. To measure the cost of counterfactuals, we follow the standard convention of using distances ($\ell_1$-norm) between factuals and counterfactuals as a proxy. For plausibility, we assess how similar counterfactuals are to observed samples in the target domain, $\mathbf{X}^+\subset\mathcal{X}^+$. We rely on the distance-based metric used in @altmeyer2024faithful,

@@

To assess our proposed training regime, we investigate the improvements in performance metrics when using CT on top of a weak baseline (BL): specifically, a multilayer perceptron (*MLP*). This is the best way to get a clear picture of how effective CT is and consistent with how assessment is done in the related literature [@ross2021learning;@teney2020learning;@goodfellow2014explaining].

@@

@tbl-main presents our main empirical findings. For all datasets except *OL* and across all test settings, the average distance of CEs from observed samples in the target class is reduced, indicating improved plausibility. The magnitude of improvements varies. For the simple synthetic datasets, distance reductions range from around 20-40% (*LS*, *Moon*) to almost 60% (*Circ*). For the real-world tabular datasets, improvements tend to be smaller but still substantial, with around 10-15% for *CH*, 11-28% for *GMSC*, 7-8% for *Cred*, and around 3% for *Adult*. For our only vision dataset (*MNIST*), distances are reduced by up to 9%. The results for our proposed divergence metric are qualitatively similar, but generally even more pronounced: for the *Circ* dataset, implausibility is reduced by almost 94% to virtually zero as we verified by the absolute outcome. Improvements for other datasets range from 28% (*Moon*) to 78% (*GMSC*). For *OL* the reduction is negative, consistent with the distance-based metric. *MNIST* is the only dataset for which the distance and divergence metrics disagree. Upon visual inspection of the image counterfactuals we find that CT clearly improves plausibility (see appendix).

@@

Key performance metrics across all datasets. **Plausibility**: Columns 2-6 show the percentage reduction in implausibility ($\text{IP}$) for varying degrees of the energy penalty used for *ECCo* at test time; column 7 shows the reduction in $\text{IP}^*$ ($\text{MMD}$), aggregated across all test specifications. **Accuracy** (columns 8-11): test accuracies and robust accuracies ($\text{Acc}^*$) for CT and the baseline (BL). **Actionability** (column 9): average reduction in costs when imposing mutability constraints.
:::

>>>>

Our key outcome of interest is improvement in explainability (Def. \ref{def-explainability}). To this end, we focus primarily on the plausibility and cost of faithfully generated counterfactuals at test time. To measure the cost, we follow the standard convention of using distances ($\ell_1$-norm) between factuals and counterfactuals as a proxy. For plausibility, we assess how similar CEs are to the observed samples in the target domain, $\mathbf{X}^+\subset\mathcal{X}^+$. We rely on the distance-based metric used in @altmeyer2024faithful,

@@

To assess CT, we investigate the improvements in performance metrics when using it on top of a weak baseline (BL): a multilayer perceptron (*MLP*). This is the best way to get a clear picture of the effectiveness of CT, and it is consistent with how assessment is done in the related literature [@goodfellow2014explaining;@ross2021learning;@teney2020learning].

@@

@tbl-main presents our main empirical findings. For all datasets except *OL* and across all test settings, the average distance of CEs from observed samples in the target class is reduced, indicating improved plausibility. The magnitude of improvements varies. For the simple synthetic datasets, distance reductions range from around 20-40% (*LS*, *Moon*) to almost 60% (*Circ*). For the real-world tabular datasets, improvements tend to be smaller but still substantial, with around 10-15% for *CH*, 11-28% for *GMSC*, 7-8% for *Cred*, and around 3% for *Adult*. For the vision dataset (*MNIST*), distances are reduced by up to 9%. The results for our proposed divergence metric are qualitatively similar, but generally even more pronounced: for the *Circ* dataset, implausibility is reduced by almost 94% to virtually zero as we verified by the absolute outcome. Improvements for other datasets range from 28% (*Moon*) up to 78% (*GMSC*). For *OL* the reduction is negative, consistent with the distance-based metric. *MNIST* is the only dataset for which the distance and divergence metrics disagree. Upon visual inspection of the image counterfactuals we find that CT clearly improves plausibility (see supplementary appendix).

@@

Key performance metrics across all datasets (column 1). **Plausibility**: Columns 2-6 show the percentage reduction in implausibility ($\text{IP}$) for varying degrees of the energy penalty $\lambda_{\text{egy}}$ used for *ECCo* at test time; column 7 shows the reduction in $\text{IP}^*$ ($\text{MMD}$), aggregated across all test specifications. **Accuracy** (columns 8-11): test accuracies and robust accuracies ($\text{Acc}^*$) for CT and the baseline (BL). **Actionability** (column 12): average reduction in costs when imposing mutability constraints reported for the four datasets for which we could identify meaningful features to protect.
:::

### 2025-03-15 paper/sections/method.qmd
Back to 16 pages, emphasize robustness a bit more

![Illustration of how CT improves model explainability: (a) conventional training, all mutable; (b) CT, all mutable; (c) conventional, *age* immutable; (d) CT, *age* immutable. The decision boundary is shown in green along with training data colored according to their ground-truth label: $y^+=2$ (orange) and $y^-=1$ (blue). Stars indicate CEs for the target class.](/paper/figures/poc.svg){#fig-poc}

>>>>

![Illustration of how CT improves model explainability: (a) conventional training, all mutable; (b) CT, all mutable; (c) conventional, *age* immutable; (d) CT, *age* immutable. The linear decision boundary is shown in green along with training data colored according to their ground-truth label: $y^-=1$ in blue and $y^+=2$ in orange. Stars indicate counterfactuals in the target class.](/paper/figures/poc.svg){#fig-poc}
